A new method to estimate dark matter halo concentrations

### 2015-09-21 paper/article.tex
minor edits in the introduction

One the most popular parameterization for a halo radial density
distribution is known as the Navarro-Frenk-White (NFW) profile
\citep{NFW}.
This profile is a  double power law in radius, where the transition break
happens at the so-called scale radius $r_s$.
The ratio between the scale radius and the virial radius, $R_v$ which
defines a natural scale for the halo, is known as the concentration
$c=R_v/r_s$. 
Simulations show that the concentration is a strong function of halo mass and
redshift.

@@

Observationally, the relationship between halo mass and concentration
could provides a potential test of LCDM on galactic scales.
With this promise a great deal of effort has been
invested in calibrating this relationship with simulations
\citep{Neto2007,Duffy2008,Munoz2011,Prada2012,Ludlow2014} but also
finding the best possible way to constraint it with observations
\citep{Buote2007,Comerford2007,Mandelbaum2008,Giocoli2014,Foex2014,Shan2015}.

From the computational point of view there are two main methods to
estimate the concentration parameter of a dark matter halo in a N-body
simulation.
The first method takes the particles composing the halos
and bins them in logarithmic radii to estimate the density in each
bin, then it proceeds with a fit of this density estimation as a
function of the radius.
A second method uses an analytic property of the
NFW that relates the maximum of the ratio of the circular velocity to
the virial velocity.
In this method the user estimates this ratio as a function of radius
to determine its maxium value.
Then, the user can take this value to find the roots of an equation
which represent the concentration parameter.

The first method is straightforward to apply, but it has two
disadvantages.
First, it requires a large number of particles in
order to have a proper density estimate in each bin.
This makes the method robust only when halos with at least  $10^3$
particles are fitted.
The second problem is that there is not a way to estimate the optimal
bin size.
Different choices affect the results of the fit.

@@

In this paper we propose a new method to estimate the dark matter halo
concentration in halos obtained in numerical simulations.
It consists in building the cumulative mass profile from the
particle data in the N-body simulation to find the best possible
concentration value using a Markov Chain Monte Carlo (MCMC)
methodology by comparing the data against the analytic expectation.

Our proposal has two advantages with respect to the two methods mentioned
above.
First, it does not involve data binning.
Second, it does not throw away data points.
Furthermore, the MCMC approach allows for an straightforward
estimation of the uncertainties in the concentration parameter, and
inclusion of informative priors.

>>>>

One the most popular parameterization for a dark matter halo radial density
distribution is the Navarro-Frenk-White (NFW) profile
\citep{NFW}.
This profile is a double power law in radius, where the transition break
happens at the so-called scale radius $r_s$.
The ratio between the scale radius and the virial radius $R_v$, which
defines a natural scale for the halo, is known as the concentration
$c=R_v/r_s$. 
Simulations show that the concentration is a strong function of halo mass and
redshift.

@@

Observationally, the relationship between halo mass and concentration
could provide a potential test of LCDM on galactic scales.
For this motivation a great deal of effort has been
invested in calibrating this relationship with simulations
\citep{Neto2007,Duffy2008,Munoz2011,Prada2012,Ludlow2014} and 
finding the best possible way to constraint it with observations
\citep{Buote2007,Comerford2007,Mandelbaum2008,Giocoli2014,Foex2014,Shan2015}.

From the computational point of view there are two main methods to
estimate the concentration parameter of a dark matter halo in a N-body
simulation.
The first method takes the particles composing the halos
and bins them in logarithmic radii to estimate the density in each
bin, then it proceeds with a fit of this density estimation as a
function of the radius.
A second method uses an analytic property of the
NFW that relates the maximum of the ratio of the circular velocity to
the virial velocity.
The concentration can be then found as the root of an algorithmnic
equation dependent on this maximum value.

The first method is straightforward to apply but presents two
disadvantages.
First, it requires a large number of particles in
order to have a proper density estimate in each bin.
This makes the method robust only for halos with at least  $10^3$
particles.
The second problem is that there is not a way to estimate the optimal
bin size, different choices produce different results for the
concentration.

@@

In this paper we propose a new method to estimate the dark matter halo
concentration in halos obtained in numerical simulations.
It builds the cumulative mass profile from the particle data in the
N-body simulation to find the best possible concentration value using
a Markov Chain Monte Carlo (MCMC) methodology.

Our proposal has two advantages with respect to the two methods mentioned
above. 
It does not involve data binning and does not throw away data points.
Furthermore, the MCMC approach allows for an straightforward
estimation of the uncertainties in the concentration parameter, and
inclusion of informative priors.
